---
title: "Analysis of Chance of Admission"
author: "Michael Dang"
format: 
  html:
    embed-resources: true
date: 11-20-2024
---

The dataset that will be used is: <https://www.kaggle.com/datasets/mohansacharya/graduate-admissions>

Where the data dictionary is found:

```{r setup}
#| message: false
#| warning: false
library(broom)
library(tidyverse)
library(readr)
```


## Load the dataset
```{r read}
dataset <- read_csv(
  file = "../data/Admission_Predict.csv")
names(dataset) <- tolower(names(dataset))
glimpse(dataset)
```


Rename the column
```{r rename}
dataset |> 
  rename(
    serial_no = "serial no.",
    gre_score = "gre score",
    toefl_score = "toefl score",
    university_rating = "university rating",
    chance_of_admit = "chance of admit"
  ) -> dataset

glimpse(dataset)
```

## Descriptive statistics
```{r group-means}
dataset |>
  group_by(university_rating) |>
  summarize(
    chance_of_admit_mn=mean(chance_of_admit),
    chance_of_admit_sd=sd(chance_of_admit),
    n=n())
```
As the University Rating increase the Chance of Admit decrease. More than half of the applications were applied to 2nd and 3rd University Rating.

## Box plot
```{r box-plot1}
dataset |> 
  ggplot(aes(x = as.factor(university_rating), y = chance_of_admit)) +
    geom_boxplot(fill = "lightblue", color = "black") +
    ggtitle("Graph drawn by Michael Dang on 2024-11-15") +
    xlab("University Rating") +
    ylab("Chance of Admit") +
    coord_flip() +
    theme_minimal()
```

There is evidence of non-normality in the distribution of 'Chance of Admit' across university ratings. The graph exhibits right skewness, especially in lower ratings, and includes outliers across all ratings, with a concentration in the 3rd and 4th ratings.

## Histogram plot 

```{r histogram1}
dataset |> 
  ggplot(aes(x = gre_score)) +
    geom_histogram(binwidth = 5, fill = "skyblue", color = "black") +
    ggtitle("Graph drawn by Michael Dang on 2024-11-15") +
    xlab("GRE Score") +
    ylab("Frequency") 
```

```{r histogram2}
dataset |> 
  ggplot(aes(x = cgpa)) +
    geom_histogram(binwidth = 0.3, fill = "lightgreen", color = "black") +
    ggtitle("Histogram of CGPA") +
    xlab("CGPA") +
    ylab("Frequency") 
```
Both histogram of GRE Score and CGPA shows roughly bell-shaped distribution, which suggests it may approximate a normal distribution.

## Hypothesis test

- Null hypothesis ($H_0$): The mean "Chance of Admit" is the same across all university ratings.

- Alternative hypothesis ($H_a$): The mean "Chance of Admit" differs for at least one university rating group.





