---
title: "Sample size justification for postural sway study"
format: 
  html:
    embed-resources: true
editor: source
---

This program evaluates various sample size calculations for a proposed replication of the postural-sway study in a different population. It was written by Steve Simon on 2024-10-08 and is placed in the public domain.

## Scenario 1

-   Replicate postural sway study
    -   Different populations
    -   Same outcome measure
-   Research hypothesis, $H_0\ \mu_1 - \mu_2 = 0$
-   Standard deviations: 9.77, 4.09
-   MCID = 4

```{r power-1}
power.t.test(
  n=NULL,
  delta=4,
  sd=9.8,
  sig.level=0.05,
  power=0.9,
  type="two.sample",
  alternative="two.sided")
```

With a sample of 128 patients, we would have 90% power for detecting a 4 unit difference in postural sway, using a two-sided test at an alpha level of 0.05.

## Scenario 2, MCID = 2

```{r power-2}
power.t.test(
  n=NULL,
  delta=2,
  sd=9.8,
  sig.level=0.05,
  power=0.9,
  type="two.sample",
  alternative="two.sided")
```

If we wanted to be able to detect a 2 unit difference, we would need a sample size that is about four times as large.
    
## Scenario 3, MCID=8

```{r power-3}
power.t.test(
  n=NULL,
  delta=8,
  sd=9.8,
  sig.level=0.05,
  power=0.9,
  type="two.sample",
  alternative="two.sided")
```
If we wanted to be able to detect a 8 unit difference, we would need a sample size that is about four times as smaller.

## Scenario 4, sd=4.9

```{r power-4}
power.t.test(
  n=NULL,
  delta=4,
  sd=4.9,
  sig.level=0.05,
  power=0.9,
  type="two.sample",
  alternative="two.sided")
```
With half of the standard deviation, if we wanted to be able to detect a 4 unit difference, we would need a sample size of 32 patients.

## Scenario 5, sd=19.6

```{r power-5}
power.t.test(
  n=NULL,
  delta=4,
  sd=19.6,
  sig.level=0.05,
  power=0.9,
  type="two.sample",
  alternative="two.sided")
```
For the double of the standard deviation, if we wanted to be able to detect a 4 unit difference, we would need a sample size of 505 patients.

## Scenario 6, alpha=0.01
```{r power-6}
power.t.test(
  n=NULL,
  delta=4,
  sd=9.8,
  sig.level=0.01,
  power=0.9,
  type="two.sample",
  alternative="two.sided")
```
For the significant level of 1% with standard deviation is 9.8, if we wanted to be able to detect a 4 unit difference, we would need a sample size of 180 patients.

## Scenario 7, alpha=0.10
```{r power-7}
power.t.test(
  n=NULL,
  delta=4,
  sd=9.8,
  sig.level=0.1,
  power=0.9,
  type="two.sample",
  alternative="two.sided")
```
For the significant level of 10% with standard deviation is 9.8, if we wanted to be able to detect a 4 unit difference, we would need a sample size of 103 patients.

## Scenario 8, power=0.8
```{r power-8}
power.t.test(
  n=NULL,
  delta=4,
  sd=9.8,
  sig.level=0.1,
  power=0.8,
  type="two.sample",
  alternative="two.sided")
```
With the power decrease by 0.1 and a 9.8 standard deviation, if we wanted to be able to detect a 4 unit difference, we would need a sample size of 75 patients.

## Scenario 9, power=0.95
```{r power-9}
power.t.test(
  n=NULL,
  delta=4,
  sd=9.8,
  sig.level=0.1,
  power=0.95,
  type="two.sample",
  alternative="two.sided")
```
With the power increase by 0.15 and a 9.8 standard deviation, if we wanted to be able to detect a 4 unit difference, we would need a sample size of 130 patients.
