---
title: "Analysis of Titanic dataset"
author: "Michael Dang"
format: 
  html:
    embed-resources: true
date: 11-20-2024
---

This program reads data on survival of passengers on the Titanic. Find more
information in the [data dictionary][web00].

[web00]: https://github.com/pmean/data/blob/main/files/titanic.yaml


## Load the tidyverse library

```{r}
#| label: setup
#| message: false
#| warning: false
library(broom)
library(epitools)
library(tidyverse)
```

#### Comments on the code

For most of your programs, you should load the [tidyverse library][web01]. The messages and warnings are suppressed.

[web01]: https://www.tidyverse.org/

In previous programs, I put a label for each chunk inside the curly braces ({}).
It is recommended instead to put the label on a separate line inside the program
chunk. It is a bit more work to provide a unique label for each chunk, but it
helps quite a bit to isolate where to look when your code produces an error.

## Read the data and view a brief summary

```{r}
#| label: read
ti <- read_tsv(
  file="../data/titanic.txt",
  col_names=TRUE,
  col_types="ccncn",
  na="NA")
names(ti) <- tolower(names(ti))
glimpse(ti)
```

#### Comments on the code

Use read_tsv from the [readr package][web02] to  read this file. Use 
col_names=TRUE because the column names are included as the first row of the 
file. The col_types="ccncn" specifies the first second and fourth columns as
strings and the third and fifth as numeric. There are missing values in this
dataset, designated by the letters "NA".

[web02]: https://readr.tidyverse.org/

## Replace numeric codes for survived

```{r}
#| label: replace-numbers
ti$survived <-
	factor(
		ti$survived,
		level=1:0,
		labels=c("yes", "no"))
```

#### Comments on the code

The [factor function][web03] places the levels of a categorical variable in a specific order and (optionally) attaches labels to each level. In this code, the
number codes are reordered so that 1 appears first followed by 0. The labels "yes" and "no" are attached to these two codes.

[web03]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/factor.html

## Question 1
Create a new variable, third_class that indicates whether a passenger is in third class or not. The code would look something like this.

```{r}
#| label: question_1
ti$third_class <- 
  case_when(
    ti$pclass == "1st" ~ "no",
    ti$pclass == "2nd" ~ "no",
    ti$pclass == "3rd" ~ "yes")
```

How many passengers were in the thrid class?

```{r}
#| label: question_1_count
sum(ti$third_class == "yes", na.rm = TRUE)
```

## Question 2
What are the probabilities of survival for third class passengers. How does this compare to the probability of survival for the other passengers.


## Get counts of third class by survival

```{r}
#| label: counts
table1 <-xtabs(~third_class+survived, data=ti)
table1
```

#### Interpretation of the output

There were 138 third class passengers survive and 573 third class passengers died.

## Get proportions for died/survived by third class

```{r}
#| label: proportions
table1 |>
  proportions("third_class")
```

#### Interpretation of the output

The proportion of first/second class passenger who died is 48%. The proportion of third class passenger who died is much higher at 80%


## Question 3
Test the hypothesis that the survival probability is different for third class passengers and the other passengers. Interpret the p-value and confidence interval.

- Null hypothesis ($H_0$): The survival probabilities for third-class passengers and other passengers are the same.
- Alternative hypothesis ($H_A$): The survival probabilities for third-class passengers and other passengers are different.

```{r}
#| label: equality-of-proportions
prop.test(table1, correct=FALSE)
```

#### Interpretation of the output
- Since the p-value is almost 0, which is less than 0.5; hence, we can reject the null hypothesis and conclude that the survival probabilities for third-class passengers and other passengers are different.
- 95% confidence interval for the difference in survival probabilities is between 27.48% and 37.36%. Since the interval does not contain 0, hence there is a significant difference in survival probabilities for third-class passengers and other passengers.

