---
title: "Analysis of fruitfly data"
author: "Michael Dang"
format: 
  html:
    embed-resources: true
date: 11-13-2024
---

This program reads data on fruit fly longevity. Find more information in the [data dictionary][dd].

[dd]: https://github.com/pmean/data/blob/master/files/fruitfly.yaml

## Load the tidyverse library

```{r setup}
#| message: false
#| warning: false
library(broom)
library(tidyverse)
```

#### Comments on the code

For most of your programs, you should load the tidyverse library. The broom 
library converts your output to a nicely arranged dataframe. The messages and
warnings are suppressed.

## List the variable names

```{r variable-list}
vlist <- c(
  "id",
  "partners",
  "type",
  "longevity",
  "thorax",
  "sleep")
```

#### Comments on the code

When a dataset does not have variables on the first line, you need to specify 
them in the code.

## Read the data and view a brief summary

```{r read}
fly <- read_fwf(
  "../data/fruitfly.txt",
  col_types="nnnnnn",
  fwf_widths(
    widths=c(2, 2, 2, 3, 5, 3),
    col_names=vlist))
glimpse(fly)
```

#### Comments on the code

The fruitfly dataset has a fixed width format (fwf). You need to specify the
columns that each variable uses.

Notice that the two categorical variables, partners and type, are actually
numbers rather than strings. To avoid having R treat these variables as if
they were continuous, use the factor function in some of the code below.

## Question 1
Create a subset of the fruitfly data by removing the age where type equals 9. Draw a clustered boxplot with sleep as the outcome and partners and type as the categorical predictors. Interpret this graph. Is there evidence of non-normality?

```{r subset}
fly |>
  filter(type != 9) -> fly_subset
```

#### Comments on the code

If you exclude the pure control group (No females), you can analyze the two
factors, partners and type individually. Partners has two category levels, 1 for
when one female was included in the cage and 8 for when eight females were 
included in the cage. Type also has two category levels, 0 for pregnant female
fly/flies and 1 for virgin fly/flies. A male fly will not mate with a pregnant
females, so you can think of this as a second level of controls. The two factors
are crossed, meaning that every possible combination of partners and type has 
outcomes measured.

## Draw boxplot of longevity by partners and type

```{r boxplots-partners-type}
#| fig.width: 6
#| fig.height: 2.5
fly_subset |>
  ggplot(aes(factor(partners), sleep, fill=factor(type))) +
    geom_boxplot() +
    xlab("Number of partners") +
    ylab("Percentage of each day sleeping") +
    ggtitle("Graphs drawn by Michael Dnag on 2024-11-13") +
    labs(fill="Female fly type") +
    scale_fill_discrete(labels=c("Pregnant", "Virgin")) +
    coord_flip()
```
"Virgin" flies tend to sleep a higher median percentage of the day, with greater variability, while "Pregnant" flies show a narrower range and lower median. The data shows evidence of non-normality due to outliers and asymmetrical distributions in both groups.


## Question 2
Calculate descriptive statistics for sleep (mean, standard deviation, and sample size) by the combination of the two categorical predictors, partners and type. Is there evidence of heterogeneity?

```{r by-partners-and-type}
#| message: false
fly_subset |>
  group_by(type, partners) |>
  summarize(
    sleep_mn=mean(sleep),
    sleep_sd=sd(sleep),
    n=n()) -> fly_means
fly_means
```
Yes, there is evidence of heterogeneity in this data.  The standard deviations for sleep percentage vary notably between groups, with values ranging from 10.74 to 19.83.

## Question 3
Draw a line graph for the mean sleep levels compared by type and partners. Is there evidence of an interaction?

```{r line-graph}
fly_means |>
  ggplot(aes(
    factor(partners), 
    sleep_mn, 
    group=factor(type),
    color=factor(type))) +
    geom_line(linewidth=2) +
	expand_limits(y=range(fly_subset$sleep)) +
  xlab("Number of partners") +
  ylab("Lifespan in days") +
  labs(color="Female fly type") +
  scale_color_discrete(labels=c("Pregnant", "Virgin"))
```
There is little evidence of a strong interaction between the number of partners and female fly type on lifespan. Both "Pregnant" and "Virgin" flies exhibit nearly parallel lines with a slight decrease in lifespan as the number of partners increases. 

## Question 4
Analyze the sleep variable using a two factor analysis of variance with an interaction. Present and interpret the analysis of variance table.

```{r anova-with-interaction}
m1 <- aov(sleep ~ factor(partners)*factor(type), data=fly_subset)
anova(m1)
```

There is no evidence for a statistically significant interaction between the number of female partners and the type of partners (pregnant or virgin) nor their interaction significantly affects the amount of sleep. This imply no substantial impact of these factors on sleep variability.

## Question 5
What factors might make you consider using a log transformation for the sleep variable? Do not run such an analysis but tell us whether you think the data would warrant such a transformation?

Ans:

- The factor that make me consider using a log transformation for sleep variable is the data distribution is skewed or if there are large differences in variability (heteroscedasticity) between groups. In the provided data, there are instances of both high and low sleep values with varying standard deviations across groups.

- I think the data should consider a log transformation, because the data is left-skewed and has extreme outliers based on the plot on Question 1.