---
title: "Analysis of fruitfly data"
author: "Michael Dang"
format: 
  html:
    embed-resources: true
date: 11-06-2024
---

This program reads data on fruit fly longevity. Find more information in the [data dictionary][dd].

[dd]: https://github.com/pmean/data/blob/master/files/fruitfly.yaml

## Load the tidyverse library

```{r setup}
#| message: false
#| warning: false
library(broom)
library(tidyverse)
```

For most of your programs, you should load the tidyverse library. The broom library converts your output to a nicely arranged dataframe. The messages and warnings are suppressed.

## List the variable names

```{r variable-list}
fn <- "https://jse.amstat.org/datasets/fruitfly.dat.txt"
vlist <- c(
  "id",
  "partners",
  "type",
  "longevity",
  "thorax",
  "sleep")
```

When a dataset does not have variables on the first line, you need to specify them in the code.

## Read the data and view a brief summary

```{r read}
fly <- read_fwf(
  "../data/fruitfly.txt",
  col_types="nnnnnn",
  fwf_widths(
    widths=c(2, 2, 2, 3, 5, 3),
    col_names=vlist))
glimpse(fly)
```

The fruitfly dataset has a fixed width format (fwf). You need to specify the 
columns that each variable uses.

## Create cage groups

```{r cage}
fly$cage <- 
  case_when(
    fly$partners==0 & fly$type==9 ~ "No females",
    fly$partners==1 & fly$type==0 ~ "One pregnant female",
    fly$partners==1 & fly$type==1 ~ "One virgin female",
    fly$partners==8 & fly$type==0 ~ "Eight pregnant females",
    fly$partners==8 & fly$type==1 ~ "Eight virgin females")
```

The five categories represent different combinations of partners and type.

## Question 1
Review the fruitfly analysis discussed in this module. There is a second variable, sleep, that might be influenced by the presence or absence of virgin or pregnant females. Compute descriptive statistics for sleep levels in each of the five groups. Interpret these statistics

```{r longevity-means}
fly |>
  group_by(cage) |>
  summarize(
    sleep_mn=mean(sleep),
    sleep_sd=sd(sleep),
    n=n())
```

The mean sleep duration is much lower for the group with eight virgin females. Standard deviations vary but are generally consistent, indicating similar levels of variability across all groups.

## Question 2
Draw a boxplot for sleep levels in each group. Interpret the boxplots.

```{r longevity-boxplot}
#| fig.width: 6
#| fig.height: 2.5
fly |>
  ggplot(aes(cage, sleep)) +
    geom_boxplot() +
    ggtitle("Graph drawn by Michael Dang on 2024-11-06") +
    xlab("Cage residents") +
    ylab("Lifespan (days)") +
    coord_flip()
```

The boxplot shows a left-skewed with many outliers

## Question 4
Based on the previous two questions, do you believe that the assumptions of analysis of variance are met. Proceed with all of the remaining questions regardless of your conclusion here.

Answer:

- Based on the box plot, seem like the assumption of normality is violate because "One virgin female" and "Eight pregnant females" have outliers and high degree of variability. 

- Also the standard deviation across groups are not consistent, hence homogenity may get violated. 

- Assume sample were collected iid hence independence is met.

Therefore, assumption of ANOVA are not met due to lack of normality and homogenity.

## Question 4
Conduct a single factor analysis of variance, using sleep as the dependent variable and cage as the categorical predictor variable. Print an analysis of variance table. Interpret the F-ratio and the p-value.

```{r longevity-one-factor-anova}
m1 <- aov(sleep ~ cage, data=fly)
tidy(m1)
```

The F-ratio is small and the p-value is large. Conclude that there is not statistical 
difference among some or all of the population mean lifespans.

## Question 5
Calculate and interpret confidence intervals using the Tukey post hoc comparisons. Which intervals include 0 and which do not. Provide a general conclusion about which groups, if any, differ from one another.

```{r Tukey post hos}
t1 <- TukeyHSD(m1, order = TRUE)
t1
```
Since all the interval contains 0, we conclude that there are no significant differences in sleep levels between any of the cage groups.

## Question 6
Conduct a Kruskal-Wallis test. Interpret your results.

```{r kw}
kruskal.test(sleep ~ cage, data=fly)
```
The Kruskal-Wallis chi-squared: 0.35, is a very low test statistics. Also the p-value is 0.98 indicates that there is no statistically significant difference in sleep levels among the different cage group.

