---
title: "Analysis of Chance of Admission"
author: "Michael Dang"
format: 
  html:
    embed-resources: true
date: 11-20-2024
---

The dataset that will be used is: <https://www.kaggle.com/datasets/mohansacharya/graduate-admissions>

Where the data dictionary is found: <https://github.com/micho0802/Bio-Stat/blob/main/Admission_Predict_data_dictionary.yaml>

## Load the library
```{r setup}
#| message: false
#| warning: false
library(broom)
library(tidyverse)
library(readr)
```


## Load the dataset
```{r read}
dataset <- read_csv(
  file = "../data/Admission_Predict.csv", show_col_types = FALSE)
names(dataset) <- tolower(names(dataset))
glimpse(dataset)
```


Rename the column
```{r rename}
dataset |> 
  rename(
    serial_no = "serial no.",
    gre_score = "gre score",
    toefl_score = "toefl score",
    university_rating = "university rating",
    chance_of_admit = "chance of admit"
  ) -> dataset

glimpse(dataset)
```

## Descriptive statistics
```{r group-means}
dataset |>
  group_by(university_rating) |>
  summarize(
    chance_of_admit_mn=mean(chance_of_admit),
    chance_of_admit_sd=sd(chance_of_admit),
    n=n())
```
As the University Rating increase the Chance of Admit decrease. More than half of the applications were applied to 2nd and 3rd University Rating.

## Box plot
```{r box-plot1}
dataset |> 
  ggplot(aes(x = as.factor(university_rating), y = chance_of_admit)) +
    geom_boxplot(fill = "lightblue", color = "black") +
    ggtitle("Graph drawn by Michael Dang on 2024-11-15") +
    xlab("University Rating") +
    ylab("Chance of Admit") +
    coord_flip() +
    theme_minimal()
```

- There is evidence of non-normality in the distribution of 'Chance of Admit' across university ratings. The graph exhibits right skewness, especially in lower ratings, and includes outliers across all ratings, with a concentration in the 3rd and 4th ratings. 
- Hence non-normality assumption maybe violated and we may assume for heterogeneity.

#### Another boxplot

```{r box-plot2}
dataset |> 
  ggplot(aes(x = as.factor(university_rating), y = cgpa)) +
    geom_boxplot(fill = "lightblue", color = "black") +
    ggtitle("Graph drawn by Michael Dang on 2024-11-15") +
    xlab("University Rating") +
    ylab("Undergraduate GPA on the scale 10") +
    coord_flip() +
    theme_minimal()
```
- The plot shows evidence of variability in GPA across university ratings. Higher university ratings are associated with higher cumulative GPAs, but the distributions are non-uniform. Outliers are present in all university ratings, with a concentrations in the 1st and 4th ratings. The data suggests some skewness, particularly in the lower ratings, indicating a diverse range of student profiles across these ratings. 
- Hence non-normality assumption maybe violated and we may assume for heterogeneity.


## Histogram plot

```{r histogram1}
dataset |> 
  ggplot(aes(x = chance_of_admit)) +
    geom_histogram(binwidth = 0.05, fill = "lightgreen", color = "black") +
    ggtitle("Graph drawn by Michael Dang on 2024-11-15") +
    xlab("Chance of Admit") +
    ylab("Frequency") 
```
The histogram of the Chance of Admit showed the distribution is right-skewed but it is roughly center at the middle.

#### Another histogram plot
```{r histogram2}
dataset |> 
  ggplot(aes(x = cgpa)) +
    geom_histogram(binwidth = 0.3, fill = "lightgreen", color = "black") +
    ggtitle("Graph drawn by Michael Dang on 2024-11-15") +
    xlab("CGPA") +
    ylab("Frequency") 
```
The histogram of CGPA shows roughly bell-shaped distribution, which suggests it may approximate a normal distribution.

## Hypothesis test

- Null hypothesis ($H_0$): The mean "Chance of Admit" is the same across all university ratings.

- Alternative hypothesis ($H_a$): The mean "Chance of Admit" differs for at least one university rating group.

- $H_0: \mu = \mu_0$ vs $H_a: \mu \neq \mu_0$ 

## One-way ANOVA

```{r anova}
m1 <- aov(chance_of_admit ~ as.factor(university_rating), data = dataset)
tidy(m1)
```
The F-ratio is large and the p-value is small. Conclude there is a differs for at least one university rating group.

Since ANOVA indicates significant differences, perform post-hoc tests (e.g., Tukey's HSD) to identify which specific groups (university ratings) differ from each other.

```{r Tukey post-hoc-test}
TukeyHSD(m1)
```
- Higher Ratings Have Higher Chance: Each higher university rating has a significantly higher "Chance of Admit" compared to lower ratings

- Largest Difference: The largest difference is between ratings 5 and 1, diff = 0.3401, showing that students with a university_rating = 5 have a substantially higher average chance of admit compared to rating = 1.

- Smallest Significant Difference: The smallest difference is between 5 and 4, diff = 0.0701, suggesting a smaller improvement in "Chance of Admit" between these higher-rated groups.

- Statistical Significance: All pairwise comparisons have p-values less than 0.05, indicating that the differences in means are statistically significant.

